Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-04-27 paper/main.tex
supplemental references

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation. As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to blank out those bases during training by using the sample weights described in Table \ref{tab:arrays}. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Section \ref{sec:training_data}.

@@

\subsection{Metrics}
We use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Table \ref{tab:detailed_results} for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

@@

We also implemented a dilated CNN (dCNN) baseline as this kind of architecture is widely used when working with DNA data as input. Details of the neural architecture search are given in Section \ref{sec:dcnn}.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates the ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be up to 200.000 base pairs long and is determined by the phylogenetic class. The concrete lengths were chosen to keep the typical average gene length roughly proportional to the length of the sequence input. For more implementation details on this see Section \ref{sec:longer} or the source code.

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see Figure \ref{fig:length_wise_bias}). It also improves the average model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. The figures in Section \ref{sec:overlapping} show the effect of overlapping for each genome, ordered by N75. We found that overlapping tends to work best if the genomes are not very fragmented and we used it for both animals and plants.

@@

Our models, however, tend to perform less well for the very smallest and largest genomes or species that are phylogenetically the furthest away from our training genomes. This is the case for both animals and plants and is visualized by the Supplemental Figures S?-?. We do not, for example, consistently predict very well on the algae as well as the non-avian reptiles. While one of the the algae were included in our training genomes, the non-avian reptiles were not. Our model also tends to show higher uncertainty around transitions of annotation classes (e.g. from UTR to CDS) due to the lack of further post-processing of the basepair wise predictions.

AUGUSTUS does outperform us rather consistently on the smallest genomes, but falls off much more with growing genome length. The difference in prediction quality is especially strong for mammals, which tend to have large genomes with very long genes, and the largest plants.

>>>>

During data generation, we only queried for the longest transcript of each gene and did not include FASTA sequences that had no functional annotation. As Geenuff checks the genomic annotations for potential errors and is able to mark those areas, we were able to blank out those bases during training by using the sample weights described in Table \ref{tab:arrays}. The vast majority of bases masked lie in the intergenic region as the most prevelant error is a missing UTR and Geenuff marks a potentially large intergenic region for it. Table \ref{tab:statistics_all} shows statistics about the masking. The corresponding statistics for the selected training genomes only can be found in Figure S6.

@@

\subsection{Metrics}
We use two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with AUGUSTUS and further explained in Section \ref{sec:augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. (See Table \ref{tab:statistics_all} for the class distribution and Tables S1 and S2 for a more in-depth report our model performance including the intergenic class). In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes.

@@

We also implemented a dilated CNN (dCNN) baseline as this kind of architecture is widely used when working with DNA data as input. Details of the neural architecture search are given in Section 6 of the Supplements.

\subsection{Inference Techniques}
We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference time than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates the ability of our model to generalize as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be up to 200.000 base pairs long and is determined by the phylogenetic class. The concrete lengths were chosen to keep the typical average gene length roughly proportional to the length of the sequence input. For more implementation details on this see Section 3 of the Supplements or the source code.

The final predictions of a single model are constructed by overlapping predictions, that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see Figure \ref{fig:length_wise_bias}). It also improves the average model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base. The figures in Section 4 of the Supplements show the effect of overlapping for each genome, ordered by N75. We found that overlapping tends to work best if the genomes are not very fragmented and we used it for both animals and plants.

@@

Our models, however, tend to perform less well for the very smallest and largest genomes or species that are phylogenetically the furthest away from our training genomes. This is the case for both animals and plants and is visualized by the Figures S1 and S3. We do not, for example, consistently predict very well on the algae as well as the non-avian reptiles. While one of the the algae were included in our training genomes, the non-avian reptiles were not. Our model also tends to show higher uncertainty around transitions of annotation classes (e.g. from UTR to CDS) due to the lack of further post-processing of the basepair wise predictions.

AUGUSTUS does outperform us rather consistently on the smallest genomes, but falls off much more the larger the genomes get. The difference in prediction quality is especially strong for mammals, which tend to have big genomes with very long genes as well as the largest plants. Figures S2 and S4 display the comparison to AUGUSTUS by phylogenetic position.
